Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-04-21 Paper/manuscript_v2/manuscript_07_01.tex
update eq 16 of sec 5.3 and related text around

As the WBC predictor values were strictly positive and highly skewed, a logarithm transformation is used. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i|f_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i|f)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$ depending on predictors $x$ and hyperparameters $\theta=(\alpha,\ell)$,
%
\begin{equation*} 
f(\bm{x}) \sim \GP(c + \bm{\beta}\bm{x}, k(\bm{x},\bm{x}', \theta)),
\end{equation*}
where $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model.
% Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ are multivariate Gaussian distributed with mean function $\mu(\cdot)$ and covariance matrix $K$, where $\mu(\bm{x}_i)=c + \bm{\beta}\bm{x}_i$ and $K_{rs}=k(\bm{x}_r,\bm{x}_s,\theta)$, with $r,s=1,\dots,n$.
The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. A scalar lengthscale is used in the multivariate covariance function assuming an isotropic function.

Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. A general mean GP function is defined for all observations, while a second GP function only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align} 
f(\bm{x}) &\sim \GP\left(c + \bm{\beta}\bm{x},\, k(\bm{x},\bm{x}', \theta_1)\right) + g(\bm{x}), \label{ch5_eq_gpml1}
\end{align}
with
%
\begin{align} \label{ch5_eq_gpml2}
g(\bm{x}) =
\left\{
	\begin{array}{ll}
		0, & \;\; \mbox{if } \;\; x_2=1,  \\
		\GP\left(0, k(\bm{x},\bm{x}', \theta_2)\right),  & \;\; \mbox{if } \;\; x_2=2 \,.
	\end{array}
\right. 
\end{align}

%\begin{align} 
%f(\bm{x}) &\sim \GP(c + \bm{\beta}\bm{x}, k(\bm{x},\bm{x}', \theta_1)) \label{ch5_eq_gpml1} \\
%g(\bm{x}) &\sim \GP(c + \bm{\beta}\bm{x}, k(\bm{x},\bm{x}', \theta_2 \,|\, x_2\!=\!2)) \label{ch5_eq_gpml2} \\ 
%f(\bm{x} \, |\, x_2\!=\!2) &= f(\bm{x} \,|\, x_2\!=\!1) + g(\bm{x}). \nonumber
%\end{align}

\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia_02.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}
Above, $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and lengthscale of the general mean GP function, and $\theta_2$ contains the hyperparameters $\alpha_2$ and $\ell_2$ which are the marginal variance and lengthscale, respectively, of a GP function specific to the male sex ($x_2=2$).

Using the HSGP approximation, the functions $f(\bm{x})$ and $g(\bm{x})$ are approximated as in equation (\ref{eq_approxf_multi}), with the $D$-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}), and the multivariate eigenfunctions $\phi_j$ and the $D$-vector of eigenvalues $\bm{\lambda}_j$ as in equations (\ref{eq_eigenfunction_multi}) and  (\ref{eq_eigenvalue_multi}), respectively.

>>>>

As the WBC predictor values were strictly positive and highly skewed, a logarithm transformation is used. Continuous predictors were normalized to have zero mean and unit standard deviation. %Survival time was normalized to have zero mean for the logarithm of time. 
We assume a log-normal observation model for the observed survival time, $t_i$, with a function of the predictors, $f(\bm{x}_i):{\rm I\!R}^4 \to {\rm I\!R}$, as the location parameter, and $\sigma$ as the Gaussian noise: 
%
\begin{equation*}
  p(t_i \mid f_i)= \LogNormal(t_i \mid f(\bm{x}_i),\sigma^2).
\end{equation*}

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i \mid f)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$. Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. The latent function $f(\bm{x})$, besides of being centered around a linear model, is composed of a general mean GP function, $h(\bm{x})$, defined for all observations, plus a second GP function, $g(\bm{x})$, that only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align*}
h(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_0) \right),  \\[1pt] 
%
g(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_1) \mid x_2=2 \right),\\[1pt] 
%
f(\bm{x}) &= c + \bm{\beta} \bm{x} + h(\bm{x}) + g(\bm{x}). 
\end{align*}
%
Above, $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model. $\theta_0$ contains the hyperparameters $\alpha_0$ and $\ell_0$ which are the marginal variance and lengthscale of the general mean GP function, and $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and lengthscale, respectively, of a GP function specific to the male sex ($x_2=2$). Scalar lengthscales, $l_0$ and $l_1$, are used in both multivariate covariance functions, assuming isotropic functions.

%with
%%
%\begin{align} \label{ch5_eq_gpml2}
%g(\bm{x}) =
%\left\{
%	\begin{array}{ll}
%		0, & \;\; \mbox{if } \;\; x_2=1,  \\
%		\GP\left(0, k(\bm{x},\bm{x}', \theta_2)\right),  & \;\; \mbox{if } \;\; x_2=2 \,.
%	\end{array}
%\right. 
%\end{align}

\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 0mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia_02.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}

Using the HSGP approximation, the functions $h(\bm{x})$ and $g(\bm{x})$ are approximated as in equation (\ref{eq_approxf_multi}), with the $D$-dimensional (with a scalar lengthscale) squared exponential spectral density $S$ as in equation (\ref{eq_specdens_inf}), and the multivariate eigenfunctions $\phi_j$ and the $D$-vector of eigenvalues $\bm{\lambda}_j$ as in equations (\ref{eq_eigenfunction_multi}) and  (\ref{eq_eigenvalue_multi}), respectively.
